---
title: "Introduction to the bara package"
author: "Robin Gradin"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
set.seed(91827)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(foreach)
library(doRNG)
```

# Batch Adjustment by Reference Alignment (BARA)

## Introduction

Data obtained using biological data acquisition techniques are often associated with batch effects.
Batch effects can arise from a large variety of sources and can sometimes have
detrimental impacts on subsequent analyses. Batch effects are especially troublesome
in prediction problems where it is common to have a fixed prediction model for classifying
subsequently acquired test samples. Thus, the training set and the test set are
often completely correlated with batch, which can have a negative impact on the
model's prediction performance.

BARA is a method developed specifically for adjusting batch effects between 
training sets and test sets in predictive settings using reference samples.
Further, the method assumes that the variance important for making accurate predictions
can be captured in a lower dimensional space defined by the singular vectors of
the training set. Using a subset of the singular vectors containing the largest 
fraction of variance, the adjustments between the training set and the test set
is made in the compressed space spanned by the vectors. 

This package is currently under construction, but the important functions
are described below.

## Main Functions

The bara method is implemented by first creating a `BaraFit` object using the
`bara_fit()` function. The function uses the training set to identify a set of
singular vectors, where the number of vectors retained are based on either 
the `ndim` or the `loss` parameters. The function returns a list of objects that 
can be used in a subsequent step for adjusting test data. The returned list 
contains the compressed input data in `fit$x`, which can be used to define a prediction model.

```{r }
library(bara)
# Generate some "training data".
x <- matrix(rnorm(4000), nrow = 40)
# Define the index of the reference samples
ref <- c(10, 30)
# Only retain 10 dimensions of x.
ndim = 10
# Create the BaraFit object
fit <- bara_fit(x = x, ref = ref, ndim = ndim, verbose = FALSE)
head(fit$x[, 1:3], n = 3)
```

The adjustment of a test set can be made by using the `bara_adjust()` method.
The function returns the adjusted test data that can be fed into the prediction model
defined by the compressed training set.

```{r }
# Generate some test data and define the test reference samples
x_test <- matrix(rnorm(2000), nrow = 20)
ref_test <- c(5, 15)
x_test_adjusted <- bara_adjust(bara_fit = fit, x = x_test, ref = ref_test)
head(x_test_adjusted[, 1:3], n = 3)
```

Often, the optimal number of dimensions to retain can be optimized by testing.
Therefore, the package also contains a function for estimating the optimal compression factor 
(the number of dimensions to retain). Currently, the optimization can only be
performed when both a training set and a test set exist. The method is implemented in the
function `bara_optimize_ext()`. To allow flexibility, the 
function requires the user to input functions for fitting a prediction model,
classifying the test set and for evaluating the performance.

```{r tidy.opts=TRUE}
# Define response levels for the training set and the test set
y <- factor(rep(c(1, 2), each = 20))  # For the training set
y_test <- factor(rep(c(1, 2), each = 10))
# Test the method with a kNN model.
library(class)
# Define the fit function f
# The function must take the inputs: x, y and ...
fit_f <- function(x, y, ...){
  # Because the kNN function takes the training set and the test
  # set as input simultaneously, the training set and the labels are 
  # just bundled in a list.
  list(
    x = x,
    y = y
  )
}
# Define the prediction function
# The required inputs are: object, x and ...
pred_f <- function(object, x, ...){
  knn(train = object$x, test = x, cl = object$y, k = 4)
}

# Finally, define the performance method by estimating accuracy.
# Required arguments are: y, pred and ...
perf_f <- function(y, pred, ...){
  mean(y == pred) * 100
}

# Now, the optimization function can be run.
opt_results <- bara_optimize_ext(
  x_train = x,
  ref_train = ref,
  y_train = y,
  x_test = x_test,
  ref_test = ref_test,
  y_test = y_test,
  fit_fun = fit_f,
  pred_fun = pred_f,
  perf_fun = perf_f,
  perf_objective = 'maximize',
  verbose = FALSE
)
opt_results$best_dim
```
